Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2024-02-19 paper/main.tex
finished first pass at updated GLMM results section

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether they are likely to answer that 
question correctly or incorrectly. We developed a statistical approach to test this claim. 
For each quiz question a participant answered, in turn, we used Equation~\ref{eqn:prop} to estimate their knowledge at the given question's embedding space coordinate based on other questions that participant answered on the same quiz. 
We repeated this for all participants, and for each of the three quizzes. 
Then, separately for each quiz, we fit a generalized linear mixed model (GLMM) with a logistic link function to explain the likelihood of correctly answering a question as a function of estimated knowledge for its embedding coordinate, while accounting for random variation among participants and questions (see \nameref{subsec:glmm}).
To assess the predictive value of the knowledge estimates, we compared each GLMMs to an analogous (i.e., nested) ``null'' model that did not consider estimated knowledge using using parametric bootstrap likelihood-ratio tests.

@@

For the initial quizzes participants took (prior to watching either lecture),
predicted knowledge tended to be low overall, and relatively
unstructured (Fig.~\ref{fig:predictions}, left column). When we held out
individual questions and predicted their knowledge at the held-out questions'
embedding coordinates, we found no reliable differences in the predictions when
the held-out question had been correctly versus incorrectly answered. This
``null'' effect persisted when we used \textit{all} of the Quiz 1 questions
from a given participant to predict a held-out question (``All questions''; $\U
= 50587,~p = 0.723$), when we used questions from one lecture to predict
knowledge at the embedding coordinate of a held-out question about the
\textit{other} lecture (``Across-lecture''; predicting knowledge for held-out
\textit{Four Fundamental Forces Questions} using \textit{Birth of Stars}
questions: $\U = 8244,~p = 0.184$; predicting knowledge for held-out
\textit{Birth of Stars} questions: $\U = 8202.5,~p = 0.161$), and when we used
questions from one lecture to predict knowledge at the embedding coordinate of
a held-out question about the \textit{same} lecture (``Within-lecture'';
\textit{Four Fundamental Forces}: $\U = 7681.5,~p = 0.746$; \textit{Birth of
Stars}: $\U = 8125,~p = 0.204$). We believe that this reflects a floor effect:
when knowledge is low everywhere, there is little signal to differentiate
between what is known versus unknown.

After watching \textit{Four Fundamental Forces}, predicted knowledge for
held-out questions that were answered correctly (from the second quiz;
Fig.~\ref{fig:predictions}, middle column) exhibited a significant positive
shift relative to held-out questions that were answered incorrectly. This held
when we included all questions in the analysis ($\U = 58332,~p < 0.001$), when
we predicted knowledge across-lectures (\textit{Four Fundamental Forces}: $\U =
6749.5,~p = 0.014$; \textit{Birth of Stars}: $\U = 8480,~p = 0.016$), and when
we predicted knowledge at the embedding coordinates of held-out \textit{Four
Fundamental Forces} questions using other \textit{Four Fundamental Forces}
questions from the same quiz and participant ($\U = 7224,~p < 0.001$). This
difference did \textit{not} hold for within-lecture knowledge predictions at
knowledge at embedding space coordinates of \textit{Birth of Stars} questions
($\U = 7419,~p = 0.739$). Again, we suggest that this might reflect a floor
effect whereby, at that point in the participants' training, their knowledge
about the content of the \textit{Birth of Stars} material is relatively low
everywhere in that region of text embedding space.

Finally, after watching \textit{Birth of Stars}, predicted knowledge for
held-out correctly answered questions (from the third quiz;
Fig.~\ref{fig:predictions}, right column) was higher than for held-out
incorrectly answered questions. This held when we included all questions in the
analysis ($\U = 38279,~p = 0.022$), when we carried out across-lecture
predictions (\textit{Four Fundamental Forces}: $\U = 6684.5,~p = 0.032$;
\textit{Birth of Stars}: $\U = 6414.5,~p = 0.002$), and when we carried out
within-lecture knowledge predictions for held-out \textit{Birth of Stars}
questions using other \textit{Birth of Stars} questions from the same quiz and
participant ($\U = 6126,~p = 0.006$). However, we found the \textit{opposite}
effect when we carried out within-lecture knowledge predictions for held-out
\textit{Four Fundamental Forces} questions using other \textit{Four Fundamental
Forces} questions from the same quiz and participant ($\U = 6734,~p = 0.027$).
Specifically, on Quiz 3, our knowledge predictions for held-out correctly
answered questions about \textit{Four Fundamental Forces} were reliably
\textit{lower} than those for their incorrectly answered counterparts.
Speculatively, we suggest that this may reflect participants forgetting some of
the \textit{Four Fundamental Forces} content. If this forgetting happens in a
relatively ``random'' way (with respect to spatial distance within the text
embedding space), then it could explain why some held-out questions about
\textit{Four Fundamental Forces} were answered incorrectly, even if questions
at nearby coordinates (i.e., about similar content) were answered correctly.
This might lead our approach to over-estimate knowledge for held-out questions
about ``forgotten'' knowledge that participants answered incorrectly. Taken
together, the results in Figure~\ref{fig:predictions} indicate that our
approach can reliably predict acquired knowledge (especially about recently
learned content), and that the knowledge predictions are generalizable across
the content areas spanned by the two lectures, while also specific enough to
distinguish between questions about more subtly different content within the
same lecture.

>>>>

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether they are likely to answer that 
question correctly or incorrectly. We developed a statistical approach to test this claim. 
For each quiz question a participant answered, in turn, we used Equation~\ref{eqn:prop} to estimate their knowledge at the given question's embedding space coordinate based on other questions that participant answered on the same quiz. 
We repeated this for all participants, and for each of the three quizzes. 
Then, separately for each quiz, we fit a generalized linear mixed model (GLMM) with a logistic link function to explain the likelihood of correctly answering a question as a function of estimated knowledge for its embedding coordinate, while accounting for random variation among participants and questions (see \nameref{subsec:glmm}).
To assess the predictive value of the knowledge estimates, we compared each GLMMs to an analogous (i.e., nested) ``null'' model that did not consider estimated knowledge using parametric bootstrap likelihood-ratio tests.

@@

%When we estimated participants' knowledge for each Quiz~1 question based on all other Quiz~1 questions, we found an inverse relationship. 
%Specifically, higher estimated knowledge at the embedding coordinate at a held-out question was associated with a lower likelihood of answering the question correctly ($\textrm{odds ratio}\ (OR) = 0.136,\ \textrm{likelihood-ratio test statistic}\ (\lambda_{LR}) = 19.749,\ \textrm{95\% CI} = [14.352,\ 26.545],\ p = 0.001$). 
%However, this inverse relationship in fact represents the expected result under our null hypothesis (that estimated knowledge is \textit{not} predictive of success on a question). 
%An intuition for this can be taken from the expected outcome of same analysis based on the simple proportion correct, rather than estimated knowledge. 
%Suppose a participant answered $n$ out of 13 quiz questions correctly. 
%If we held out a single correctly answered question and computed the proportion of remaining questions answered correctly, that proportion would be $(n - 1) / 12$. 
%Whereas if we held out a single incorrectly answered question, the proportion of remaining questions answered correctly would be $n / 12$.

In performing this set of analyses, our null hypothesis is that the knowledge estimates we compute based on the quiz questions' embedding coordinates do \textit{not} provide useful information about participants' abilities to answer those questions. 
What result might we expect to see if this is the case? 
To provide an intuition for this, consider the expected outcome if we carried out these same analyses using a simple proportion-correct measure in lieu of our knowledge estimates. 
Suppose a participant correctly answered $n$ out of 13 questions on a given quiz. 
If we held out a single correctly answered question and computed the proportion of remaining questions answered correctly, that proportion would be $(n - 1) / 12$. 
Whereas if we held out a single \textit{incorrectly} answered question and did the same, that proportion would be $n / 12$. 
Thus for a given participant and quiz, a ``knowledge estimate'' computed as the simple (i.e., unweighted) remaining proportion-correct is perfectly inversely related to success on a held-out question: it will always be \textit{lower} for correctly answered questions than for incorrectly answered questions. 
Given that our knowledge estimates are computed as a weighted version of this same proportion-correct score (where each held-in question's weight reflects its embedding-space distance from the held-out question; see Eqn.~\ref{eqn:prop}), if these weights are uninformative (e.g., simply randomly distributed), then we should expect to see this same inverse relationship emerge, on average. 
It is only if the spatial relationships among the quiz questions' embedding coordinates map onto participants' knowledge in a meaningful way that we would we expect this relationship to be non-negative [\textbf{PHRASING}].

When we fit a GLMM to estimates of participants' knowledge for each Quiz~1 question based on all other Quiz~1 questions, we observed this null-hypothesized inverse relationship. 
Specifically, higher estimated knowledge at the embedding coordinate of a held-out Quiz~1 question was associated with a lower likelihood of answering the question correctly (odds ratio $(OR) = 0.136$, likelihood-ratio test statistic $(\lambda_{LR}) = 19.749$, 95\%\ $\textnormal{CI} = [14.352,\ 26.545],\ p = 0.001$). 
However, when we repeated this analysis for quizzes 2 and 3, the direction of this relationship reversed: higher estimated knowledge for a given question predicted a greater likelihood of answering it correctly (Quiz~2: $OR = 2.905,\ \lambda_{LR} = 17.333,\ 95\%\ \textnormal{CI} = [14.966,\ 29.309],\ p = 0.002$; Quiz~3: $OR = 3.238,\ \lambda_{LR} = 6.882,\ 95\%\ \textnormal{CI} = [6.228,\ 8.184],\ p = 0.017$). 
Taken together, these results suggest that our knowledge estimations can reliably predict participants' likelihood of success on individual quiz questions, provided they have at least some amount of structured knowledge about the underlying concepts being tested. 
In other words, when participants' correct responses primarily arise from knowledge about the content probed by each question (e.g., after watching one or both lectures), these successes can be predicted from their ability to answer other questions about conceptually similar content (as captured by embedding-space distance).
However, when a sufficiently large portion of participants' correct responses (presumably) reflect successful random guessing (such as on a multiple-choice quiz taken before viewing either lecture), our approach fails to accurately predict these successes since they do not map onto embedding space distances in a meaningful way [\textbf{PHRASING}].

We observed a similar pattern when we fit GLMMs to estimates of participants' knowledge for each question about one lecture derived from other questions about the \textit{same} lecture. 
Specifically, for questions that participants answered on Quiz~1, prior to watching either lecture, knowledge for the embedding coordinates of \textit{Four Fundamental Forces}-related questions estimated using other \textit{Four Fundamental Forces}-related questions did not reliably predict whether those questions were answered correctly ($OR = 1.891,\ \lambda_{LR} = 2.293,\ 95\%\ \textnormal{CI} = [2.091,\ 2.622],\ p = 0.139$). 
The same was true of knowledge estimates for \textit{Birth of Stars}-related questions based on other \textit{Birth of Stars}-related questions ($OR = 0.722,\ \lambda_{LR} = 5.115,\ 95\%\ \textnormal{CI} = [0.094,\ 0.146],\ p = 0.738$). 
When we recomputed these within-lecture knowledge estimates using questions from Quiz~2---which participants took immediately after viewing \textit{Four Fundamental Forces} but prior to viewing \textit{Birth of Stars}---we found that they now reliably predicted success on \textit{Four Fundamental Forces}-related questions ($OR = 9.023,\ \lambda_{LR} = 18.707,\ 95\%\ \textnormal{CI} = [10.877,\ 22.222],\ p = 0.001$) but not on \textit{Birth of Stars}-related questions ($OR = 0.306,\ \lambda_{LR} = 5.115,\ 95\%\ \textnormal{CI} = [4.624,\ 5.655],\ p = 0.055$).
Using participants' responses from Quiz~3 (taken immediately after viewing \textit{Birth of Stars}), we found that within-lecture knowledge estimates for \textit{Birth of Stars}-related questions could now reliably predict success on those questions ($OR = 5.467,\ \lambda_{LR} = 10.670,\ 95\%\ \textnormal{CI} = [7.998, 12.532],\ p = 0.006$). 
However, within-lecture knowledge estimates for \textit{Four Fundamental Forces} questions answered on Quiz~3 were no longer directly related to the likelihood of successfully answering them and instead exhibited the inverse relationship we would expect to arise from unstructured knowledge (with respect to the embedding space; $OR = 0.013,\ \lambda_{LR} = 14.648,\ 95\%\ \textnormal{CI} = [10.695, 23.096],\ p = 0.001$).
Speculatively, we suggest that this may reflect participants forgetting some of the \textit{Four Fundamental Forces} content (e.g., perhaps in favor of prioritizing encoding the just-watched \textit{Birth of Stars} content in preparation for the third quiz). 
If this forgetting happens in a relatively ``random'' way (with respect to spatial distance within the embedding space), then it could explain why some held-out questions about \textit{Four Fundamental Forces} were answered incorrectly, even if questions at nearby coordinates (i.e., about similar content) were answered correctly.
This might lead our approach to over-estimate knowledge for held-out questions about ``forgotten'' knowledge that participants answered incorrectly.
% ALTERNATE EXPLANATION -- embedding space is essentially ``saturated'' with correctly answered questions, so just like how on quiz 1 when relatively few questions are correct, most questions ``around'' them will be incorrect, on quiz 3 when relatively few questions are incorrect, most questions nearby will be correct. And because of this, on average, when the ``held-out'' question is one of the few incorrect ones, there will tend to be more correct ones ``held in'' than there will be when the held-out question is correct.
% Also, maybe worth noting: while negative relationship is significant, it's super weak -- per the model, a "1-unit" increase in estimated knowledge corresponds to only a 1.28% decrease in probability of correct answer (p = OR / (1 + OR)). For comparison, for quiz3/within-lecture/birth of stars, 1-unit increase in estimated knowledge corresponds to a 84.5% increase in probability. So decrease is sig. but basically negligible.
Taken together, these results suggest that our approach can distinguish between questions about different content covered by a single lecture when participants have sufficiently structured knowledge about that lecture's content, though this specificity may decrease further in time from when the lecture in question was viewed.

Finally, when we fit GLMMs to estimates of participants' knowledge for questions about one lecture based on questions (from the same quiz) about the other lecture, we observed a similar but slightly more nuanced pattern. 
Essentially, while the previous set of analyses suggest that our approach's ability to make \textit{specific} predictions within content areas depends on participants having a minimum level of knowledge about the given content, the across-lecture analyses we performed suggest that our ability to \textit{generalize} these predictions across different content areas requires that participants' level of knowledge about the content used to make predictions be reasonably similar to their level of knowledge about the content for which these predictions are made [\textbf{PHRASING}].
We found that using questions answered on Quiz~1, participants abilities to correctly answer questions about \textit{Four Fundamental Forces} could be predicted from their responses to questions about \textit{Birth of Stars} ($OR = 1.896,\ \lambda_{LR} = 7.205,\ 95\%\ \textnormal{CI} = [6.224, 7.524],\ p = 0.039$) and their ability to correctly answer \textit{Birth of Stars}-related questions could be predicted from their responses to \textit{Four Fundamental Forces}-related questions ($OR = 1.522,\ \lambda_{LR} =  6.448,\ 95\%\ \textnormal{CI} = [5.656, 6.843],\ p = 0.043$).
We note, however, that these Quiz~1 knowledge estimates suffer from the same ``noise'' due to the (presumably) higher rate of participants successfully guessing correct answers on Quiz~1 as noted above, and as a result provide the weakest signal of any of the knowledge estimates that we found to reliably predict success.
When we repeated this analysis using questions from Quiz~2, we found participants' responses to \textit{Four Fundamental Forces}-related questions did not reliably predict their success on \textit{Birth of Stars}-related questions ($OR = 1.865,\ \lambda_{LR} = 3.205,\ 95\%\ \textnormal{CI} = [3.027, 3.600],\ p = 0.125$), nor did their responses to \textit{Birth of Stars}-related questions reliably predict their success on \textit{Four Fundamental Forces}-related questions ($OR = 3.490,\ \lambda_{LR} = 3.266,\ 95\%\ \textnormal{CI} = [3.033, 3.866],\ p = 0.094$). 
\textbf{Sentence about why this makes sense given that participants hadn't viewed BoS yet. i.e., when predicting held-out FFF questions, correct vs. incorrect labels for held-in q's aren't meaningfully structured w.r.t. embedding space; when predicting held-out BoS q's, whether or not held-out q was correctly answered isn't meaningfully related to spatial structure of correctly answered q's in embedding space.}
However, when we again computed these across-lecture knowledge predictions using questions from Quiz~3 (when participants had now viewed \textit{both} lectures, we found that we could again reliably predict success on questions about \textit{Four Fundamental Forces} ($OR = 11.294),\ \lambda_{LR} = 11.055,\ 95\%\ \textnormal{CI} = [9.126, 18.476],\ p = 0.004$) and \textit{Birth of Stars} ($OR = 7.302),\ \lambda_{LR} = 7.068,\ 95\%\ \textnormal{CI} = [6.490, 8.584],\ p = 0.017$).
Across all three versions of these analyses, our results suggest that our knowledge estimations can reliably predict participants' abilities to answer individual quiz questions, distinguish between questions about similar content, and generalize across content areas, provided that participants' quiz responses reflect a minimum level of ``real'' knowledge about both content on which these predictions are based and that for which they are made [\textbf{PHRASING}].

% our approach works when participants have a minimal baseline level of knowledge about content predicted and used to predict
% our approach generalizes when knowledge of content used to predict can be assumed to be a reasonable indicator of knowledge of content predicted
% our approach has enough specificity to distinguish between content within the same lecture when it was just watched -- maybe when people forget a little bit they forget "randomly"?.

% potential new transition/motivation -- in the previous analyses, we identified a particular set of constraints on our estimates of participants' knowledge. This made us wonder about another potential constraint: how far away in topic space does the relevance of being able to answer a question extend and influence ability to answer a different question?
